---
title: 'Machine Learning: Exercise Prediction'
author: "Julie Foster"
date: "November 3, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Project:  Exercise Prediction

## Introduction
One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants.

### Background
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

## Data
The training data for this project are available here:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv
The test data are available here:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

### Download and load the data:

```{r}
if(!file.exists("./data")){dir.create("./data")}
fileUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
download.file(fileUrl,destfile="./data/pml-training.csv")

fileUrl2 <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
download.file(fileUrl2,destfile="./data/pml-testing.csv")

pmltraining <- read.csv("./data/pml-training.csv")
pmltesting <- read.csv("./data/pml-testing.csv")
```

### Data Exploration:
```{r}
names(pmltraining)

library(caret)
```

### Cleaning the data

After looking at the data, it looks like removing columns that don't contribute to the accelerometer measurements is the first step:
```{r}
colremove <-  grepl("^X|timestamp|window", names(pmltraining))
training <- pmltraining[, !colremove]
trainCleaned <- training[, sapply(training, is.numeric)]
testing <- pmltesting[, !colremove]
testCleaned <- testing[, sapply(testing, is.numeric)]
```

The next step in getting the data ready for model fitting is to remove columns with missing data:
```{r}
sum(complete.cases(trainCleaned))
trainData <- trainCleaned[, colSums(is.na(trainCleaned)) == 0]
testData <- testCleaned[, colSums(is.na(testCleaned)) == 0]
```
The "testData" set will be used in the final step in this analysis.      
     
Adding in the "classe" column back into the training data set:
```{r}
trainData$classe <- pmltraining$classe
```


## Cross Validation: 
Then, we can split the cleaned training set into a training data set (70%) and a new testing data set (30%). 
```{r}
set.seed(7) 
inTrain <- createDataPartition(trainData$classe, p=0.70, list=F)
training <- trainData[inTrain, ]
testing <- trainData[-inTrain, ]
```


## Finding a model
Trying to find an appropriate model, beginning with random forest:
```{r cache=TRUE}
control <- trainControl(method="cv", number=3)
modelRf <- train(classe~., data=training, method="rf", 
                 trControl=control)
modelRf
modelRf$finalModel
```

The accuracy results from this model (98.9%), are exceptionally good, so this will be the model to be used in the predictions on the testing data set.

## Predictions
```{r}
predRf <- predict(modelRf, newdata=testing)
```

## Error comparisons

out-of-sample error:
```{r}
confusionMatrix(predRf, testing$classe)
```
Both the accuracy and the Cohen’s kappa indicator of concordance indicate that the predictor seems to have a low out of sample error rate.

## Final Step - Apply to the original test set:
```{r}
predRf_test <- predict(modelRf, newdata=testData)
predRf_test
```


